Competing Bandits: \\ The Perils of Exploration under Competition

### 2019-02-07 paper/content/introduction.tex
Minor edits.

\gaedit{We ask whether algorithms that are traditionally considered ``better" in the literature on multi-armed bandits are incentivized to be deployed under competition.} We investigate this issue via extensive numerical experiments in a stylized duopoly model. In our model, two firms compete for users and simultaneously learn from them. Each firm commits to a multi-armed bandit algorithm, and \emph{explores} according to this algorithm. Users select between the two firms based on the current reputation score: rewards from the firm's algorithm, averaged over a recent time window. Each firm's objective is to maximize its \gaedit{expected} market share (the fraction of users choosing this firm).

@@

We consider a \emph{permanent duopoly} in which both firms start at the same time, as well as \emph{temporary monopoly}: a duopoly with a \gaedit{first-mover}. \swcomment{confused about how they correspond to ``competition intensity''}Accordingly, the intensity of competition in the model varies from ``permanent monopoly" (just one firm) to ``\gadelete{early entry}\gaedit{incumbent}" (the first-mover in temporary monopoly) to permanent duopoly to ``\gadelete{late entry}\gaedit{entrant" (late-arriver in temporary monopoly)}. \gaedit{We find that, even in the permanent duopoly, in equilibrium competition incentivizes firms to play a ``greedy algorithm" that does not explicitly explore and even more so if the firm is a late arriver in a market.} This algorithm also prevails under monopoly, simply because it tends to be easier to deploy. \gaedit{Whereas the incumbent in the temporary monopoly is incentivized to deploy a more advanced exploration algorithm that performs better in the long run. As a result, we find that consumer welfare is highest under temporary monopoly and that even increasing the number of firms in the market past a duopoly makes consumers weakly worse off.}

Interpreting the adoption of better algorithms as ``innovation", our findings can be framed in terms of an ``inverted-U relationship" between competition and innovation (see Figure~\ref{fig:inverted-U}). This relationship -- too little or too much competition is bad for innovation, but intermediate levels of competition tend to be better -- is a familiar theme in the economics literature, dating back to \cite{Schumpeter-42}. \swedit{However, the ``inverted-U relationship'' is driven by different aspects in our model than the ones in the existing literature in economics. In our case, the barriers for innovations arise entirely from the reputational consequences of exploration in competition, as opposed to the R\&D costs (which is the more standard cause in prior work)}.

\gaedit{The findings on temporary monopoly shed light on the ``first-mover advantage" phenomenon in the digital economy. Being first in the market in the digital economy gives firms free data to learn from (a ``data advantage") as well as a more definite, and possibly better reputation compared to an entrant (a ``reputation advantage").}

\gaedit{Beyond the welfare implications, we also investigate the degree to which the incumbent's data advantage serves as a barrier to entry. We run additional experiments that attempt to isolate the impact of the data advantage and reputation advantage. We find that the data advantage from the incumbent's ``small head start'' can lead to a significant incumbency advantage, especially when the incumbent commits to a more advanced bandit algorithm. Our results show that even a small degree of data asymmetry can be amplified to be a larger market advantage through competition dynamics. This effect is more pronounced when the incumbent commits to a more advanced bandit algorithm since the incumbent does not have to suffer from the reputational consequences of exploration while the entrant does. Our results shed lights on the importance of considering competition dynamics when reasoning about the anti-trust question of whether data can become a significant barrier to entry in an industry.}
\gadelete{
\swedit{Beyond the welfare implications, we also investigate the barrier to entry that follows from the incumbent's data advantage.  We run additional experiments that attempt to isolate the impact of the data advantage and reputation advantage. We find that the data advantage from the incumbent's ``small head start'' can lead to a significant incumbency advantage, especially when the incumbent commits to a more advanced bandit algorithm. Our results show that even a small degree of data asymmetry can be amplified to be a larger market advantage through competition dynamics. \gadelete{This is in contrast to the data advantage in a static environment such as discussed in \cite{varian2018artificial,bajari2018impact}, in which a small amount of additional data can only provide limited improvement in the predictive accuracy.} The effect is stronger when the incumbent commits to a more advanced bandit algorithm due to

>>>>

We ask whether algorithms that are traditionally considered ``better" in the literature on multi-armed bandits are incentivized to be deployed under competition. We investigate this issue via extensive numerical experiments in a stylized duopoly model.

\xhdr{Our model.} Two firms compete for users and simultaneously learn from them. Each firm commits to a multi-armed bandit algorithm, and \emph{explores} according to this algorithm. Users select between the two firms based on the current reputation score: rewards from the firm's algorithm, averaged over a recent time window. Each firm's objective is to maximize its \gaedit{expected} market share (the fraction of users choosing this firm).

@@

We consider a \emph{permanent duopoly} in which both firms start at the same time, as well as \emph{temporary monopoly}: a duopoly with a first-mover. Accordingly, the intensity of competition in the model varies from ``permanent monopoly" (just one firm) to ``incumbent" (the first-mover in temporary monopoly) to permanent duopoly to ``entrant" (late-arriver in temporary monopoly).

\xhdr{Main findings.}
We find that, even in the permanent duopoly, in equilibrium competition incentivizes firms to play a ``greedy algorithm" that does not explicitly explore and even more so if the firm is a late arriver in a market. This algorithm also prevails under monopoly, simply because it tends to be easier to deploy. Whereas the incumbent in the temporary monopoly is incentivized to deploy a more advanced exploration algorithm that performs better in the long run. As a result, we find that consumer welfare is highest under temporary monopoly and that even increasing the number of firms in the market past a duopoly makes consumers weakly worse off.

Interpreting the adoption of better algorithms as ``innovation", our findings can be framed in terms of an ``inverted-U relationship" between competition and innovation (see Figure~\ref{fig:inverted-U}). This relationship -- too little or too much competition is bad for innovation, but intermediate levels of competition tend to be better -- is a familiar theme in the economics literature, dating back to \cite{Schumpeter-42}. However, the ``inverted-U relationship'' is driven by different aspects in our model than the ones in the existing literature in economics. In our case, the barriers for innovations arise entirely from the reputational consequences of exploration in competition, as opposed to the R\&D costs (which is the more standard cause in prior work).

\xhdr{Additional findings.}
The findings on temporary monopoly shed light on the ``first-mover advantage" phenomenon in the digital economy. Being first in the market in the digital economy gives firms free data to learn from (a ``data advantage") as well as a more definite, and possibly better reputation compared to an entrant (a ``reputation advantage").

Beyond the welfare implications, we also investigate the degree to which the incumbent's data advantage serves as a barrier to entry. We run additional experiments that attempt to isolate the impact of the data advantage and reputation advantage. We find that the data advantage from the incumbent's ``small head start'' can lead to a significant incumbency advantage, especially when the incumbent commits to a more advanced bandit algorithm. Our results show that even a small degree of data asymmetry can be amplified to be a larger market advantage through competition dynamics. This effect is more pronounced when the incumbent commits to a more advanced bandit algorithm since the incumbent does not have to suffer from the reputational consequences of exploration while the entrant does. Our results shed lights on the importance of considering competition dynamics when reasoning about the anti-trust question of whether data can become a significant barrier to entry in an industry.

\gadelete{
\swedit{Beyond the welfare implications, we also investigate the barrier to entry that follows from the incumbent's data advantage.  We run additional experiments that attempt to isolate the impact of the data advantage and reputation advantage. We find that the data advantage from the incumbent's ``small head start'' can lead to a significant incumbency advantage, especially when the incumbent commits to a more advanced bandit algorithm. Our results show that even a small degree of data asymmetry can be amplified to be a larger market advantage through competition dynamics. \gadelete{This is in contrast to the data advantage in a static environment such as discussed in \cite{varian2018artificial,bajari2018impact}, in which a small amount of additional data can only provide limited improvement in the predictive accuracy.} The effect is stronger when the incumbent commits to a more advanced bandit algorithm due to
